---
title: "Transmission Type has no Impact on Car Gas Mileage"
author: "Tim Wise"
date: "January 2016"
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document:
    fig_caption: yes
    keep_md: yes
---

```{r, echo=FALSE, error=TRUE, message=FALSE, warning=FALSE}
library(car)
library(corrgram)
library(graphics)
library(gvlma)
library(lattice)
library(knitr)
library(MASS)
library(scales)
library(stats)
```


```{r, eval=TRUE, echo=FALSE}
# Create D the working data set
# . add mpg to gpm
# . add hpwt 
# . turn transmission type into factor

data(mtcars)

gpm  <- (1 / mtcars$mpg) * 100
hpwt <- mtcars$hp / mtcars$wt
D    <- cbind(gpm, hpwt, mtcars[,!names(mtcars) %in% c('mpg')])

D[mtcars$am == 0,'trans'] <- 'Automatic'
D[mtcars$am == 1,'trans'] <- 'Manual'
D[,'trans'] <- as.factor(D[,'trans'])
#str(D)
```


```{r, gpmByTransFit, eval=TRUE, echo=FALSE}
fit.gpm.vs.trans <- lm(gpm ~ trans, data=D)
```

```{r, eval=TRUE, echo=FALSE}
betas <- coef(summary(fit.gpm.vs.trans))
cis   <- confint(fit.gpm.vs.trans)

atrans.gpm.avg      <- betas['(Intercept)', 'Estimate']
atrans.gpm.ci.lower <-   cis['(Intercept)',  '2.5 %']
atrans.gpm.ci.upper <-   cis['(Intercept)', '97.5 %']
atrans.gpm.error    <- abs(atrans.gpm.avg - atrans.gpm.ci.lower)

mtrans.gpm.avg.delta      <- betas['transManual', 'Estimate']
mtrans.gpm.ci.lower.delta <-   cis['transManual',  '2.5 %']
mtrans.gpm.ci.upper.delta <-   cis['transManual', '97.5 %']

mtrans.gpm.avg      <- atrans.gpm.avg      + mtrans.gpm.avg.delta
mtrans.gpm.ci.lower <- atrans.gpm.ci.lower + mtrans.gpm.ci.lower.delta  
mtrans.gpm.ci.upper <- atrans.gpm.ci.upper + mtrans.gpm.ci.upper.delta   
mtrans.gpm.error    <- abs(mtrans.gpm.avg  - mtrans.gpm.ci.lower)

atrans.pvalue <- betas["(Intercept)", "Pr(>|t|)"]
mtrans.pvalue <- betas["transManual", "Pr(>|t|)"]

adj.r.square <- summary(fit.gpm.vs.trans)$adj.r.squared
```


### Executive Summary

We looked at the Motor Trend cars data set to see if  the type of 
transmission in a car had a significant effect on its fuel consumption, 
and if so, to quantify that effect.

At first glance, we found that cars with manual transmissions 
had better fuel economy, consuming on average
`r round(mtrans.gpm.avg, 1)` gallons per 100 miles (gpm) 
(+/- `r round(mtrans.gpm.error, 1)`)
compared to an average of 
`r round(atrans.gpm.avg, 1)` gpm 
(+/- `r round(atrans.gpm.error, 1)`)
for cars with automatic transmissions.
*But correllation is not causation* and just because there is a difference
in fuel economy doesn't mean transmission type is the cause of the
difference.

Digging deeper, we found that a car's fuel consumption was primarily a function 
of its weight and that the type of transmission was not significant.
The reason why cars with manual transmission had better mileage was because
in this data set the transmission type and vehicle weight were confounders 
and were not independent. 
Given a car's weight you could reasonably predict its transmission type --
The high mileage cars were light and had manual transmissions, 
while the low mileage cars were heavy and had automatic transmissions. 


### Analysis Details

We did two transformations on the cars data set. First, we inverted and scaled
*miles per gallon* (mpg) to get *gallons per 100 miles* (gpm), or kilomiles.
It is a better metric for comparing the fuel consumption among cars (see ...). 
Also, we added a column, dividing horsepower (hp) by weight (wt) to get a 
the *horsepower per 1000 lbs*. 
As we will see, this gives us a metric that is a measure of power that is not 
correlated with weight.

```{r, eval=TRUE, echo=FALSE}
atrans.gpm.median <- median(D[D$am == 0, 'gpm'])
mtrans.gpm.median <- median(D[D$am == 1, 'gpm'])
```

First, let's compare the fuel consumption of cars with automatic transmissions 
to those  with  manual transmissions, see **Figure 1**. 
There is a sizable difference in the median mileage, with 
manual transmissions consuming
`r round(mtrans.gpm.median, 2)` gpm compared to
`r round(atrans.gpm.median, 2)` gpm for automatics.

Next, let's test whether the difference is significant by doing an 
intercept-only regression of fuel consumption as a function of transmission 
type, see **Figure 2**. We see:

- In the Coefficients, the p-value (Pr(>|t|) for manual transmissions (transManaul) is
`r round(mtrans.pvalue, 5)` and less than 0.05. That means
the difference between automatics and manuals is indeed significant.

- In the Coefficients, the Estimate column says that automatics (Intercept) consume 
on average 
`r round(atrans.gpm.avg,       2)` gpm, and manuals (transManual) consume
`r round(mtrans.gpm.avg.delta, 2)` gpm less, for an average of
`r round(mtrans.gpm.avg,       1)` gpm.

- In the Confint, we see the 95th percentile interval for gpm, with automatics 
(Intercept) consuming
`r round(atrans.gpm.ci.lower, 2)` to 
`r round(atrans.gpm.ci.upper, 2)` gpm and manuals (transManual) consuming 
`r round(mtrans.gpm.ci.lower.delta, 2)` to 
`r round(mtrans.gpm.ci.upper.delta, 2)` gpm *less*, for an interval of
`r round(mtrans.gpm.ci.lower, 2)` to
`r round(mtrans.gpm.ci.upper, 2)` gpm.


```{r, gpmCorrelation, eval=TRUE, echo=FALSE}
corr.matrix <- round(cor(D[,!names(D) %in% c('trans')]), 2)
corr.matrix[lower.tri(corr.matrix,)] <- ""
```

Let's investigate further looking at what variables affect fuel consumption the most. 
**Figure 3** shows the correlation between variables in the data set. We see:   

- In the first row,  fuel consumption (gpm) is correlated strongest to weight (wt) 
(`r corr.matrix['gpm', 'wt']`).   

- Weight (wt) is correlated to transmission type (am)
(`r corr.matrix['wt', 'am']`).
This is a hint that weight and tranmission may be confounders.

- Weight (wt) is strongly correlated to number of cylinders (cyl), 
displacement (disp), horsepower, but not horsepower/weight (hpwt).

Let's explore the relationship between fuel consumption and weight, 
see **Figure 4**. Here we plainly see that cars with the lowest fuel
consumption are the lightest and have manual transmissions, and 
the cars with the highest fuel consumption are the heaviest and have 
automatic transmissions.

Let's build a multivariate regression starting with fuel consumption as a 
function of weight and see if transmission type is significant to the equation. 
We consider potential regressors in a forward stepwise fashion, using 
the function  `add1()` to help decide which regressors to add. 
`add1()` produces an F-statistic and a p-value (like an `anova()`
comparison) for each possible regressor. We look for p-values < 0.05 and high
F values.

**Figure 5** show the output of `add1()` given the base model (gpm ~ wt).
Horsepower/weight (hpwt) has the most influence, with horsepower (hp) and 
quarter mile time (qsec) next in line. From those three variables, we choose 
to add hpwt  because 1) it is less correlated to weight than horsepower, so 
we avoid  multiliearity among regressors, and 2) it is a cleaner definition 
of power than quarter mile time which is a mish mash of power, drag, wind 
resistance, etc.


```{r, gpmByWtHpwtFit, eval=TRUE, echo=FALSE}
fit.gpm.vs.wt.hpwt <- lm(gpm ~ wt + hpwt, data=D)
```  

```{r, eval=TRUE, echo=FALSE}
betas <- coef(summary(fit.gpm.vs.wt.hpwt))
cis   <- confint(fit.gpm.vs.wt.hpwt)

intercept.avg      <- betas['(Intercept)', 'Estimate']
intercept.ci.lower <- cis['(Intercept)',  '2.5 %']
intercept.ci.upper <- cis['(Intercept)', '97.5 %']

wt.avg.delta      <- betas['wt', 'Estimate']
wt.ci.lower.delta <-   cis['wt',  '2.5 %']
wt.ci.upper.delta <-   cis['wt', '97.5 %']

hpwt.avg.delta      <- betas['hpwt', 'Estimate']
hpwt.ci.lower.delta <-   cis['hpwt',  '2.5 %']
hpwt.ci.upper.delta <-   cis['hpwt', '97.5 %']

wt.pvalue   <- betas["wt",   "Pr(>|t|)"]
hpwt.pvalue <- betas["hpwt", "Pr(>|t|)"]

adj.r.square <- summary(fit.gpm.vs.wt.hpwt)$adj.r.squared
```


We add hpwt to the model. Running `add1()` against the new model shows no 
more variables can improve the model.  **Figure 6** shows the summary of the 
final fit.


\newpage
\newpage

### Appendix  

This section contains the figures referenced in the Analysis Details.  

---

```{r, eval=TRUE, echo=TRUE}
boxplot(gpm ~ trans, data=D, 
        main="Mileage by Transmission Type", ylab="Gallons per Kilomile")
```  

**Figure 1**: Cars with manual transmissions use less fuel 
than those with automatic transmissions. But is the difference significant?  

---

```{r, gpmByTransFit, eval=FALSE, echo=TRUE}
```  
```{r, eval=TRUE, echo=TRUE}
summary(fit.gpm.vs.trans)$coef; round(confint(fit.gpm.vs.trans), 2)
```  

**Figure 2**: A regression fit of mpg ~ trans shows the difference between
automatic and manual transmissions is statistically 
significant (Pr(>|t|) = `r round(mtrans.pvalue, 5)` < 0.05)
and that manuals consume on average 
`r round(mtrans.gpm.avg.delta, 2)` less gpm than automatics, with a range of 
`r round(mtrans.gpm.ci.lower.delta, 2)` to 
`r round(mtrans.gpm.ci.upper.delta, 2)` gpm less.

---

```{r, gpmCorrelation, eval=TRUE, echo=TRUE}
```
```{r}
kable(as.data.frame(corr.matrix), digits=2, align='r', format='markdown')
```

**Figure 3**: Fuel consumption, the first row, is correlated most to weight 
(`r corr.matrix['gpm', 'wt']`) . It is also 
correlated to number of cylinders, displacment, and horsepower, all of which 
are highly correlated amongst themselves. Note that weight and horsepower/weight
are not correlated (`r corr.matrix['wt', 'hpwt']`).

---

```{r, eval=TRUE, echo=TRUE}
scatterplot(gpm ~ wt | trans, data=D,
            main="Mileage vs Weight by Transmission Type", 
            ylab="Gallons per Kilomile", xlab="Weight (Klbs)", 
            labels=row.names(D), smoother=FALSE, legend.coords='bottomright')
```

**Figure 4**: Here we plainly see that cars with the lowest fuel
consumption are the lightest and have manual transmissions, and 
the cars with the highest fuel consumption are the heaviest and have 
automatic transmissions.

---

```{r, eval=TRUE, echo=TRUE}
fit.gpm.vs.wt <- lm(gpm ~ wt, data=D);
add1(fit.gpm.vs.wt, D, test="F")
```  

**Figure 5**: With a base model of fuel consumption as a function of weight, 
use `add1()` to evaluate with variable to add next. Horsepower/weight 
(hpwt) has the highest F value and would have the most influence.

---

```{r, gpmByWtHpwtfit, eval=TRUE, echo=TRUE}
fit.gpm.vs.wt.hpwt <- lm(gpm ~ wt + hpwt, data=D)
```  
```{r, eval=TRUE, echo=TRUE}
summary(fit.gpm.vs.wt.hpwt)
round(confint (fit.gpm.vs.wt.hpwt),  2)
round(sqrt(vif(fit.gpm.vs.wt.hpwt)), 4)
shapiro.test(rstandard(fit.gpm.vs.wt.hpwt))
```  

**Figure 6**: The summary of the final model. Both regressors are significant.
The Adjusted R-squared value says the model accounts for 
`r percent(adj.r.square)` of the variance of the fuel consumption.
The variance inflation factors (vif) are near 1 and indicate there is no
multicolliniarity among the regressors. The Shapiro-Wilk test, with a 
p-value > 0.05, says that the residuals of the fit are normally distributed.

---


```{r eval=TRUE, echo=FALSE, fig.width=6, fig.height=3}
par(mfrow=c(1,3))
plot(fit.gpm.vs.wt.hpwt, which=1)
qqPlot(fit.gpm.vs.wt.hpwt, id.n=0, main="Normal Q-Q")
hist(rstandard(fit.gpm.vs.wt.hpwt), col='gray',  main='Standard Residuals', xlab='')
par(mfrow=c(1,1))
```  

**Figure 7**: The diagnostic plots for gpm ~ wt + hpwt. In Residual vs Fitted,
the values look reasonably randomly distributed about 0. In Normal Q-Q, the 
points lie reasonably along the diagonal. And the histogram of standardized
residuals looks like a normal distribution. I am satisfied with this fit.
