---
title: "Transmission Type has no Impact on Car Fuel Consumption"
author: "Tim Wise"
date: "January 2016"
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document:
    fig_caption: yes
    keep_md: yes
---

```{r, echo=FALSE, error=TRUE, message=FALSE, warning=FALSE}
library(car)
library(corrgram)
library(graphics)
library(gvlma)
library(lattice)
library(knitr)
library(MASS)
library(scales)
library(stats)
```


```{r, eval=TRUE, echo=FALSE}
# Create D the working data set
# . add mpg to gpm
# . add hpwt 
# . turn transmission type into factor

data(mtcars)

gpm  <- (1 / mtcars$mpg) * 100
hpwt <- mtcars$hp / mtcars$wt
D    <- cbind(gpm, hpwt, mtcars[,!names(mtcars) %in% c('mpg')])

D[mtcars$am == 0,'trans'] <- 'Automatic'
D[mtcars$am == 1,'trans'] <- 'Manual'
D[,'trans'] <- as.factor(D[,'trans'])
#str(D)
```


```{r, gpmByTransFit, eval=TRUE, echo=FALSE}
fit.gpm.vs.trans <- lm(gpm ~ trans, data=D)
```

```{r, eval=TRUE, echo=FALSE}
betas <- coef(summary(fit.gpm.vs.trans))
cis   <- confint(fit.gpm.vs.trans)

atrans.gpm.avg      <- betas['(Intercept)', 'Estimate']
atrans.gpm.ci.lower <-   cis['(Intercept)',  '2.5 %']
atrans.gpm.ci.upper <-   cis['(Intercept)', '97.5 %']
atrans.gpm.error    <- abs(atrans.gpm.avg - atrans.gpm.ci.lower)

mtrans.gpm.avg.delta      <- betas['transManual', 'Estimate']
mtrans.gpm.ci.lower.delta <-   cis['transManual',  '2.5 %']
mtrans.gpm.ci.upper.delta <-   cis['transManual', '97.5 %']

mtrans.gpm.avg      <- atrans.gpm.avg + mtrans.gpm.avg.delta
mtrans.gpm.ci.lower <- atrans.gpm.avg + mtrans.gpm.ci.lower.delta  
mtrans.gpm.ci.upper <- atrans.gpm.avg + mtrans.gpm.ci.upper.delta   
mtrans.gpm.error    <- abs(mtrans.gpm.avg  - mtrans.gpm.ci.lower)

atrans.pvalue <- betas["(Intercept)", "Pr(>|t|)"]
mtrans.pvalue <- betas["transManual", "Pr(>|t|)"]

adj.r.square <- summary(fit.gpm.vs.trans)$adj.r.squared
```


### Executive Summary

We looked at the Motor Trend cars data set to see if  the type of 
transmission in a car had a significant effect on its fuel consumption, 
and if so, to quantify that effect.

At first glance, we found that cars with manual transmissions 
used less fuel, consuming on average
`r round(mtrans.gpm.avg, 1)` gallons per 100 miles (gpm) 
(+/- `r round(mtrans.gpm.error, 1)`)
compared to an average of 
`r round(atrans.gpm.avg, 1)` gpm 
(+/- `r round(atrans.gpm.error, 1)`)
for cars with automatic transmissions.
*But correllation is not causation* and just because there is a difference
in fuel consumption doesn't mean transmission type is the cause of the
difference.

Digging deeper, we found that a car's fuel consumption was primarily a function 
of its weight and that the type of transmission was, in the end, not significant.
The reason why cars with manual transmissions used less fuel was because
in this data set the transmission type and vehicle weight were confounders 
and were not independent. 
The high mileage cars were light and had manual transmissions, 
while the low mileage cars were heavy and had automatic transmissions. 


### Analysis Details

We did two transformations on the cars data set. First, we inverted and scaled
miles per gallon (mpg) to get gallons per 100 miles.
*Gallons per kilomile (gpm)* is a better metric for comparing the fuel consumption among cars. 
Also, we added a column, dividing horsepower (hp) by weight (wt) to get a 
the *horsepower per 1000 lbs*. 
As we will see, this gives us a a measure of power that is not 
correlated with weight.

```{r, eval=TRUE, echo=FALSE}
atrans.gpm.median <- median(D[D$am == 0, 'gpm'])
mtrans.gpm.median <- median(D[D$am == 1, 'gpm'])
```

First let's compare the fuel consumption of cars with automatic transmissions 
to those with  manual transmissions.
**Figure 1a** shows there is a sizable difference in the fuel consumption of 
manuals verses automatics. 
We verify the difference is significant by doing an  intercept-only
regression of fuel consumption as a function of transmission type.
See **Figure 2** for an interpretation of the regression results. 
In short, manuals use less fuel, consuming on average
`r round(mtrans.gpm.avg, 1)` gpm 
(+/- `r round(mtrans.gpm.error, 1)`)
compared to an average of 
`r round(atrans.gpm.avg, 1)` gpm 
(+/- `r round(atrans.gpm.error, 1)`)
for automatics.


```{r, gpmCorrelation, eval=TRUE, echo=FALSE}
corr.matrix <- round(cor(D[,!names(D) %in% c('trans')]), 2)
corr.matrix[lower.tri(corr.matrix,)] <- ""
```

Digging deeper, in **Figure 3**, we see that fuel consumption is most 
correlated to weight.
Plotting fuel consumption verses weight in **Figure 4**, 
we see transmission type is not spread evenly over weight.
(We see another view of that relationship in **Figure 1b**.)
This reason for the significant difference in fuel consumption 
between automatics and manuals.
We also see (looking at the two trend lines) that if we account for weight, 
there doesn't seem to be much difference in between the fuel consumption of 
automatics and manuals. 
Let's build a regression model to investigate that.

Starting with a base model of fuel consumption as a function of weight,
we will iterate in a forward step-wise fashion, using 
the function  `add1()` to help decide which regressors to include. 
`add1()` produces an F-statistic and a p-value (like an `anova()`
comparison) for each possible regressor. 
We look for p-values < 0.05 and high F values.

**Figure 5** shows the output of `add1()` given the base model (gpm ~ wt).
Horsepower/weight (hpwt) has the highest F value, with horsepower (hp) and 
quarter mile time (qsec) next in line. From those three variables, we choose 
to add hpwt because 
1) it is not correlated to weight like horsepower, so  we avoid multicollinearity 
among regressors, and 
2) it is design-time parameter
and a cleaner definition  of power than quarter mile time, which can only be 
measured after a car built and is a mash of power, drag, wind resistance, 
driver, etc.

```{r, gpmByWtHpwtFit, eval=TRUE, echo=FALSE}
fit.gpm.vs.wt.hpwt <- lm(gpm ~ wt + hpwt, data=D)
```  

```{r, eval=TRUE, echo=FALSE}
betas <- coef(summary(fit.gpm.vs.wt.hpwt))
cis   <- confint(fit.gpm.vs.wt.hpwt)

intercept.avg      <- betas['(Intercept)', 'Estimate']
intercept.ci.lower <- cis['(Intercept)',  '2.5 %']
intercept.ci.upper <- cis['(Intercept)', '97.5 %']

wt.avg.delta      <- betas['wt', 'Estimate']
wt.ci.lower.delta <-   cis['wt',  '2.5 %']
wt.ci.upper.delta <-   cis['wt', '97.5 %']

hpwt.avg.delta      <- betas['hpwt', 'Estimate']
hpwt.ci.lower.delta <-   cis['hpwt',  '2.5 %']
hpwt.ci.upper.delta <-   cis['hpwt', '97.5 %']

wt.pvalue   <- betas["wt",   "Pr(>|t|)"]
hpwt.pvalue <- betas["hpwt", "Pr(>|t|)"]

adj.r.square <- summary(fit.gpm.vs.wt.hpwt)$adj.r.squared
```

Adding add hpwt to the model and running `add1()` against the new model shows no 
more variables can improve the model. **Figure 6** shows a summary 
of the final fit. 
**Figure 7** shows diagnostic plots and a discussion of them.
This seems a good, parsimonious model.

Now, to answer the question of whether transmission type affects fuel 
consumption. 
In **Figure 8**, we use `anova()` and find that transmission type
would not be a significant factor in our model. So we conclude that transmission
type does not impact fuel consumption.

### Appendix  

This section contains the figures referenced in the Analysis Details.  

---

```{r, eval=TRUE, echo=FALSE, fig.width=7}
atrans.gpm.median <- median(D[D$am == 0, 'gpm'])
mtrans.gpm.median <- median(D[D$am == 1, 'gpm'])

atrans.wt.median  <- median(D[D$am == 0, 'wt'])
mtrans.wt.median  <- median(D[D$am == 1, 'wt'])

par(mfrow=c(1,2))
boxplot(gpm  ~ trans, data=D, xlab="Figure 1a", ylab="Gallons per Kilomile", main="Fuel Consumption")
boxplot(wt   ~ trans, data=D, xlab="Figure 1b", ylab="Kilopounds",           main="Weight")
par(mfrow=c(1,1))
```

**Figures 1a and 1b**: Cars with manual transmissions use less fuel 
(`r round(mtrans.gpm.median, 2)` gpm) than automatics 
(`r round(atrans.gpm.median, 2)` gpm).
and are lighter 
(`r round(mtrans.wt.median, 2)` Klbs) than automatics 
(`r round(atrans.wt.median, 2)` Klbs).
But are the differences significant?

---

```{r, gpmByTransFit, eval=FALSE, echo=TRUE}
```  
```{r, eval=TRUE, echo=TRUE}
summary(fit.gpm.vs.trans)$coef; 
round(confint(fit.gpm.vs.trans), 2)
```  

**Figure 2**: An intercept-only regression fit of mpg ~ trans shows the difference 
between automatic and manual transmissions is statistically 
significant (for transManual, Pr(>|t|) = `r round(mtrans.pvalue, 5)` < 0.05).
Automatics consume on average 
`r round(atrans.gpm.avg, 2)` gpm, 
with a confidence interval of 
`r round(atrans.gpm.ci.lower, 2)` to 
`r round(atrans.gpm.ci.upper, 2)` gpm.
Manuals consume on average 
`r round(mtrans.gpm.avg.delta, 2)` less gpm than automatics, 
with a confidence interval of 
`r round(mtrans.gpm.ci.lower.delta, 2)` to 
`r round(mtrans.gpm.ci.upper.delta, 2)` gpm less. 
Adding the deltas for manuals with the 
`r round(atrans.gpm.avg, 2)`
average for automatics gives, for manuals, 
an average of
`r round(mtrans.gpm.avg, 2)` gpm, 
with a confidence interval of 
`r round(mtrans.gpm.ci.lower, 2)` to 
`r round(mtrans.gpm.ci.upper, 2)` gpm.

---

```{r, gpmCorrelation, eval=FALSE, echo=FALSE}
```
```{r, eval=TRUE, echo=FALSE}
kable(as.data.frame(corr.matrix), digits=2, align='r', format='markdown')
```

**Figure 3**: Fuel consumption (gpm), the first row, is correlated most to weight (wt)
(`r corr.matrix['gpm', 'wt']`). 
Weight is also  correlated to displacement (disp), number of cylinders (cyl), 
and horsepower (hp), all of which  are also correlated with fuel consumption. 
Weight is not correlated to horsepower/weight (hpwt)
(`r corr.matrix['hpwt', 'wt']`).

---

```{r, eval=TRUE, echo=FALSE, fig.width=7}
scatterplot(gpm ~ wt | trans, data=D,
            main="Mileage vs Weight by Transmission Type", 
            ylab="Gallons per Kilomile", xlab="Weight (Klbs)", 
            labels=row.names(D), smoother=FALSE, legend.coords='bottomright')
```

**Figure 4**: Here we plainly see transmission type is not spread evenly across
weight. The cars with the lowest fuel consumption are the lightest and have 
manual transmissions, and the cars with the highest fuel consumption are 
the heaviest and have  automatic transmissions. It also looks like once we 
account for weight, there is not that much difference in fuel consumption
between manuals and automatics. 

---

```{r, eval=TRUE, echo=TRUE}
fit.gpm.vs.wt <- lm(gpm ~ wt, data=D);
add1(fit.gpm.vs.wt, D, test="F")
```  

**Figure 5**: With a base model of fuel consumption as a function of weight, 
use `add1()` to evaluate which variable to add next. Horsepower/weight 
(hpwt) has the highest F value and would have the most influence.

---

```{r, gpmByWtHpwtfit, eval=TRUE, echo=TRUE}
fit.gpm.vs.wt.hpwt <- lm(gpm ~ wt + hpwt, data=D)
```  
```{r, eval=TRUE, echo=TRUE}
summary(fit.gpm.vs.wt.hpwt)
round(confint (fit.gpm.vs.wt.hpwt),  2)
round(sqrt(vif(fit.gpm.vs.wt.hpwt)), 4)
shapiro.test(rstandard(fit.gpm.vs.wt.hpwt))
```  

**Figure 6**: The summary of the final model. Both regressors are significant.
The Adjusted R-squared value says the model accounts for 
`r percent(adj.r.square)` of the variance of the fuel consumption.
The variance inflation factors (vif) are near 1 and indicate there is no
multicollinearity  among the regressors. The Shapiro-Wilk test, with a 
p-value > 0.05, says that the residuals of the fit are normally distributed.
Note that the Intercept term is not significant, it's p-value is > 0.05. 
That means the reported Intercept value is not significantly different from 0, 
which makes sense for this regression.

---


```{r eval=TRUE, echo=FALSE, fig.width=8, fig.height=4}
par(mfrow=c(1,3))
plot(fit.gpm.vs.wt.hpwt, which=1)
qqPlot(fit.gpm.vs.wt.hpwt, id.n=0, main="Normal Q-Q")
hist(rstandard(fit.gpm.vs.wt.hpwt), col='gray',  main='Standard Residuals', xlab='')
par(mfrow=c(1,1))
```  

**Figure 7**: The diagnostic plots for gpm ~ wt + hpwt. In Residual vs Fitted,
the values look reasonably randomly distributed about 0. In Normal Q-Q, the 
points lie reasonably along the diagonal. And the histogram of standardized
residuals looks very much like a normal distribution. 

---

```{r eval=TRUE, echo=FALSE}
anova(lm(gpm ~ wt + hpwt, D),  lm(gpm ~ wt + hpwt + trans, D))
```  

**Figure 8**: Using `anova()`, we compare our base model (Model 1) to one that 
includes transmission type (Model 2). The Pr(>F) for Model 2 is greater than 
0.05 which says transmission type would not make a significant difference 
to the model.

---

A note on formatting: To get to five pages, I knitr'd to a Word docx file, 
then in Word, changed margins, reduced overall font size, and inserted 
page breaks to prevent figures from spanning pages.